---
title: "Analysis of population and economic impact of severe weather events in the United States."
output: html_document
---

This is the synopsis, blah blah blah.

## Data Processing ##

```{r}
library(dplyr)
library(ggplot2)
library(reshape2)
library(stringr)
```

```{r, cache=TRUE}
weather_events <- read.csv("StormData.csv")
```

The event types in the raw data do not appear to be well standardized. We first
organize event types that appear to be related. We mainly focus on event types
that have the most population and economic effects. The 
[Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
was used as a reference for standard event types.

```{r, cache=TRUE}
weather_events$EVTYPE <- 
  gsub("^TORN.*", "TORNADO", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^THUNDER.*TORM.*", "THUNDERSTORM WINDS", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^TSTM WIND.*", "THUNDERSTORM WINDS", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^FLOOD.*", "FLOOD", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^LIGHTNING.*", "LIGHTNING", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub(".*ICE STORM.*", "ICE STORM", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^WINTER STORM.*", "WINTER STORM", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^HIGH WIND.*", "HIGH WINDS", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^HURRICANE.*", "HURRICANE", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub("^WILD.*FIRE.*", "WILDFIRE", weather_events$EVTYPE)
weather_events$EVTYPE <- 
  gsub(".*FOG.*", "FOG", weather_events$EVTYPE)
```

The property and crop damage data is stored in four columns with separate 
columns for a base value and a "multiplier". We compute single values for 
property and crop damage by multiplying the base value by the multiplier.
While most data has reasonable multipliers (k=thousands, m=millions, b=billions),
some of the data has less obvious multipliers like '+', '-' and numeric values.
We chose to disregard data with unexpected multipliers as it comprises a very
minor portion of the data set and should not have any impact on the results.

```{r, cache=TRUE}
convert_multiplier <- function(mult) {
  if (mult %in% c("k", "K")) {
    return(10^3)
  } else if(mult %in% c("m", "M")) {
    return(10^6)
  } else if (mult %in% c("b", "B")) {
    return(10^9)
  } else {
    return(0)
  }
}

weather_events$PROPDMGMULT <- 
  sapply(weather_events$PROPDMGEXP, convert_multiplier)
weather_events$CROPDMGMULT <- 
  sapply(weather_events$CROPDMGEXP, convert_multiplier)
weather_events <- weather_events %>% 
  mutate(PROPDMG = PROPDMG * PROPDMGMULT, CROPDMG = CROPDMG * CROPDMGMULT)
```

## Casualties ##

We first look at severe weather events by their impact on the population. We
rank the events by casualties (injuries plus fatalities).

```{r, fig.width=12, fig.height=6}
worst_casualty_events <- weather_events %>%
  filter(FATALITIES > 0 | INJURIES > 0) %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_CASUALTIES = sum(FATALITIES + INJURIES), 
            TOTAL_FATALITIES = sum(FATALITIES),
            TOTAL_INJURIES = sum(INJURIES)) %>%
  arrange(desc(TOTAL_CASUALTIES))

top_n <- 15
sp <- split(worst_casualty_events, 
            c(rep(0, top_n), rep(2, nrow(worst_casualty_events) - top_n)))

top_worst_events <- sp[[1]]
other_worst_events <- sp[[2]] %>%
  mutate(EVTYPE="OTHER") %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_CASUALTIES = sum(TOTAL_CASUALTIES), 
            TOTAL_FATALITIES = sum(TOTAL_FATALITIES),
            TOTAL_INJURIES = sum(TOTAL_INJURIES))
  
worst_events_summary <- rbind(top_worst_events, other_worst_events)
worst_events_summary$EVTYPE <- str_wrap(worst_events_summary$EVTYPE, width=10)
worst_events_summary$EVTYPE <- 
  factor(worst_events_summary$EVTYPE, levels=worst_events_summary$EVTYPE)

molten_events <- melt(
  select(worst_events_summary, EVTYPE, TOTAL_FATALITIES, TOTAL_INJURIES), 
  id=c("EVTYPE"))
molten_events$variable <- 
  factor(molten_events$variable, 
         levels=c("TOTAL_FATALITIES", "TOTAL_INJURIES"), 
         labels=c("fatalities", "injuries"))

ggplot(molten_events, aes(EVTYPE, value, fill = variable)) +
  geom_bar(stat = "identity") +
  ggtitle("Top severe weather events by casualties") +
  xlab("") + 
  ylab("casualties") +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) 
```

### Damage ###

Damage analysis was performed by ranking the severe weather events by total 
damage (property plus crop).

```{r, fig.width=12, fig.height=6}
worst_damage_events <- weather_events %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_DAMAGE = sum(PROPDMG + CROPDMG) / 10^9, 
            TOTAL_PROPERTY_DAMAGE = sum(PROPDMG) / 10^9,
            TOTAL_CROP_DAMAGE = sum(CROPDMG) / 10^9) %>%
  arrange(desc(TOTAL_DAMAGE))

top_n <- 15
sp <- split(worst_damage_events, 
            c(rep(0, top_n), rep(2, nrow(worst_damage_events) - top_n)))

top_worst_events <- sp[[1]]
other_worst_events <- sp[[2]] %>%
  mutate(EVTYPE="OTHER") %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_DAMAGE = sum(TOTAL_DAMAGE), 
            TOTAL_PROPERTY_DAMAGE = sum(TOTAL_PROPERTY_DAMAGE),
            TOTAL_CROP_DAMAGE = sum(TOTAL_CROP_DAMAGE))

worst_events_summary <- rbind(top_worst_events, other_worst_events)
worst_events_summary$EVTYPE <- str_wrap(worst_events_summary$EVTYPE, width=10)
worst_events_summary$EVTYPE <- 
  factor(worst_events_summary$EVTYPE, levels=worst_events_summary$EVTYPE)

molten_events <- melt(
  select(worst_events_summary, 
         EVTYPE, TOTAL_PROPERTY_DAMAGE, TOTAL_CROP_DAMAGE), 
  id=c("EVTYPE"))
molten_events$variable <- 
  factor(molten_events$variable, 
         levels=c("TOTAL_PROPERTY_DAMAGE", "TOTAL_CROP_DAMAGE"), 
         labels=c("property", "crop"))

ggplot(molten_events, aes(EVTYPE, value, fill = variable)) +
  geom_bar(stat = "identity") +
  ggtitle("Top severe weather events by damage") +
  xlab("") + 
  ylab("damage (billions)") +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) 
```

## Results ##
